Pool Model Paper

### 2023-12-05 pool_model.tex
Spatial limitation ref

- See Papers.

- Different processes in bacterial communities. 
Differents states, lag-phase, growth, stationary state. What processes behind.
Activation/inhibition/waste production, resource competition, back-flow to lag-phase.

- About known models for Predictive Microbiology Modelling, e.g., Baranyi/Roberts

@@

\section{General Mathemetical Formulation}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, the system can be described with the Ordinary Differential Equations (ODEs) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
   \label{eq:model_ODE}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector descibing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentartion of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.

@@

Then the ODEs describing such a model can be written as 
\begin{equation}
\begin{cases}
    \dot{L} = -(\mu + \lambda) L\\
    \dot{G} = \lambda L + \alpha G - \mu' G\\
    \dot{D} = \mu  L + \mu' G- \beta D  
\end{cases}
\end{equation}
or in matrix form it shows the linear dependency
\begin{equation}
    \dot \mbx  = \begin{pmatrix}
                    -(\mu + \lambda) & 0             & 0      \\
                    \lambda          & \alpha - \mu' & 0      \\
                    \mu              & \mu'          & -\beta 
                 \end{pmatrix} \mbx .
\end{equation}

@@

However, except the inhibition due to resource competition, bacteria developed different others active competition mechanisms.
One of the examples of such mechanisms is a production of the antimicrobial compounds, inhibitors or toxins~\ref{wloch-salamon_effect_2008, chao_structured_1981}.

@@

\cmtb{Maybe as motivation can be spatial competitive mechanisms as adhesion to survaces and motility. 
So how much including these effects will affect system.}

\cmt{\cite{htuson_bacteriasurface_2013}: Attachment of bacteria to surfaces. (One of the mechanisms to stick to the areas with nutritients)}

To test the spatial limitations of the described above approach consider the system consisting from two bacterial species thet can be described bt two internal states: lag and growth states. 
Both of these species grow using limiting resource pool $R$ with different growth constants $\alpha$ and different lengths of lag phase determined by parameter $\lambda$.
Moreover, the species 1 is able to release a certain antimicrobial component $I$ that shows inhibitory effect on species 2 (amensalism).
The ODEs the bacterial population can be written as: 
\begin{equation}
    \begin{cases}
        \dot{L_1} = -\lambda_1 R L_1\\
        \dot{G_1} = \lambda_1 R L_1 + \alpha_1 R G_1\\
        \dot{L_2} = -\lambda_2 R L_2\\
        \dot{G_2} = \lambda_2 R L_2 + \frac{\alpha_2}{1 + \mu_I I} R G_2\\
    \end{cases}.
    \label{eq:spatial_limit_F}
\end{equation}

>>>>

Comprehensive reviews about competition/cooperation bacterial mechanisms: \cite{ghoul_ecology_2016, stubbendieck_bacterial_2016, hibbing_bacterial_2010, west_social_2007}

@@

\section{General Mathemetical Formulation}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, the system can be described with the Ordinary Differential Equations (\acp{ode}) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
   \label{eq:model_ODE}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector descibing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentartion of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.

@@

Then the \acp{ode} describing such a model can be written as 
\begin{equation}
\begin{cases}
    \dot{L} = -(\mu + \lambda) L\\
    \dot{G} = \lambda L + \alpha G - \mu' G\\
    \dot{D} = \mu  L + \mu' G- \beta D  
\end{cases}
\end{equation}
or in matrix form it shows the linear dependency
\begin{equation}
    \dot \mbx  = \begin{pmatrix}
                    -(\mu + \lambda) & 0             & 0      \\
                    \lambda          & \alpha - \mu' & 0      \\
                    \mu              & \mu'          & -\beta 
                 \end{pmatrix} \mbx .
\end{equation}

@@

However, except the inhibition due to resource competition, bacteria developed different others active competition mechanisms.
One of the examples of such mechanisms is a production of the antimicrobial compounds, inhibitors or toxins~\cite{wloch-salamon_effect_2008, chao_structured_1981}.

@@

As described above, the proposed above model operates with the total number of the bacterial counts for different species and states in the system not accounting for their spatial distributions.
Assuming that the bacteria is uniformly distributed we can simplify the problem to the system of the \acp{ode}.
However, in reality spatial effects play significant role for population survival forcing some species develop such competitive mechanism as cell adhesion to surfaces in the areas rich with nutritients~\cite{htuson_bacteriasurface_2013}.
Another mechanism is the motility of bacteria, e.g., the flagellates bacteria as \textit{E. coli} or \textit{B. subtilis} can experience the chemostatic motion of driven by oxygen concentrations~\cite{decoene_microscopic_2011}.
The bacterial social interactions such as releasing the toxins or public goods can also significantly affect the spatial structure of the community~\cite{blanchard_bacterial_2015}.
Therefore, it is rather important to establish the limitations for the model application.\\

- \cite{gorochowski_bsim_2012, li_nufeb_2019}: examples of another toolboxes for bacterial communities modelling (ABM).\\
More about the modelling of the motile bacteria can be seen in works:~\cite{decoene_microscopic_2011, rosser_modelling_2014, sokolov_physical_2012, li_amplified_2008}.

In general, bacterial communities in 2D can experience collective motion called swarming~\cite{wu_collective_2015}.
Some works modelled 2D motile bacteria:
\begin{itemize}
    \item \cite{decoene_microscopic_2011}: 2D chemostatic self-propelled bacterial motion of the flagellated bactreia (e.g. E. coli, B. subtilis).
    Chemostaxis here was driven by the oxygen concentrations.
    \item \cite{rosser_modelling_2014}: modelling of the bacterial reorintation, rotation.
    \item \cite{sokolov_physical_2012, li_amplified_2008}: modelling of the bacteria using hydrodynamic interactions and Brownian motion.
    Fluctuations in velocity and rotations.
    (\cite{li_amplified_2008}: Brownian motion gives right motion observed using microscopy).
    \item \cite{farrell_mechanically_2013}: 2D modelling using Newtonian dynamics and only mechanical interactions + nutritient diffusion.
    Branching parameter defines diff limits where diffusion dominant and limits growth, and where it's irrelevant.
    Mechanical pressure (due to bacteria-bacteria interactions) affects growth, can limit it. (!!!)
\end{itemize}

To test the spatial limitations of the described above approach consider the system consisting from two bacterial species thet can be described bt two internal states: lag and growth states. 
Both of these species grow using limiting resource pool $R$ with different growth constants $\alpha$ and different lengths of lag phase determined by parameter $\lambda$.
Moreover, the species 1 is able to release a certain antimicrobial component $I$ that shows inhibitory effect on species 2 (amensalism).
The \acp{ode} the bacterial population can be written as: 
\begin{equation}
    \begin{cases}
        \dot{L_1} = -\lambda_1 R L_1\\
        \dot{G_1} = \lambda_1 R L_1 + \alpha_1 R G_1\\
        \dot{L_2} = -\lambda_2 R L_2\\
        \dot{G_2} = \lambda_2 R L_2 + \frac{\alpha_2}{1 + \mu_I I} R G_2\\
    \end{cases}.
    \label{eq:spatial_limit_F}
\end{equation}
